### 2012-09-14 paper/chemokine_main.tex
Near final draft, waiting for author summary and acknowledgements edits

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.  T-cell navigation to the infected region is undirected but focused migration is made possible by local cytokine and chemokine production from infected epithelial cells.  We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using simulations in a purpose-built agent-based mathematical model (CyCells).

Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.  Of the eight chemokines tested, only secretion of CXCL10 (IP-10) and CCL5 (RANTES) was stimulated by infection with these three influenza strains, with IP-10 dominating the directed migration due to higher concentrations.  A delayed differential equation model was fit to the empirical per cell chemokine production rates, coupled with published T cell migration parameters, to calibrate the spatially explicit model in order to test inter-strain variation on T cell recruitment.  T cells appear to be sufficiently sensitive to chemokine to function as well as they possibly can.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production and large plaque size.  The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.  Expanding plaque sizes isolate infected cells, impeding T-cell discovery and subsequent clearance of the infection over time.

@@

@@

Mathematical models have used detailed mouse data to characterize the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.  Such whole-response modeling has considerable potential to inform strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.  Models using ordinary differential equations to predict temporal events, however, do not account for spatial constraints, and must make assumptions about the efficiency of the spread and interaction of populations.

Here we study the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).  To develop a model with tractable simulations, some components of host defense are abstracted and the model therefore does not aim to predict the true outcome for each strain examined.   We ask how very small foci of infected tissue can attract limited numbers of activated CD8 T cells after release into the systemic vascular system.  The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.   A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.

@@

The standard hypothesis posits that the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.  Our model confirms this hypothesis, but we also observed three interesting characteristics of chemokine-directed migration not revealed by earlier data.  First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.  Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at increasing concentrations above the threshold.  Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.

@@

Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions. The lung was represented as a two-dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node. At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.

@@

In the model, epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}. \emph{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from {incubating} to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for 17 hours and then die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells. Apoptoic cells continue to secrete virus until they die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T-cells have three states. \emph{Searching} T-Cells begin emerging from the lymph node at five days post infection. \textit{Searching} T-Cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine. \emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{searching} and are reintroduced to a new random location in the lung. When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection. \emph{Chemotaxing} T-Cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells. \emph{Chemotaxing} cells decay exponentially with an average lifespan of two hours.

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \emph{expressing} epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration. Chemokine diffuses across the tissue but has no
direct effect beyond activating T-Cells. Both particle types decay exponentially.

@@

\begin{table}
\begin{center}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion in Airway & $.0318 \mu m^2/s$ & \cite{Beauchemin2006} \\
  Viral Decay in Airway &  $1/day$ & \cite{Lee2009} \\
  Chemokine Diffusion Rate & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
  Chemokine Decay Rate &  $3.8508\cdot10^{-4}$/s & Selected$^1$\\
  Infection Sensitivity Rate &  $2$ \textit{hour/virion} &  Selected$^2$ \\
  Incubation Time &  $10$ \textit{hours} & \cite{Mitchell2011} \\
  Expression Time &  $16.7$ \textit{hours} & \cite{Mitchell2011} \\
  Apoptosis Time & $1$ \textit{hour} & \cite{Ganusov2008} \\
  T-Cell Production Rate & $777$\textit{/hour} & \cite{Miao2010} \\ 
  Blood Circulation Time & $6$ \textit{min} & \cite{Banerjee2010b} \\
  Search Time In Chemokine-Free Tissue & $10$ \textit{min} & \cite{Banerjee2010b} \\
  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Speed (Chemotaxis) & $3 \mu m/s$ & \cite{Miller2003} \\
  T-Cell Sensitivity to Chemokine & $10 ng/mL$ & \cite{Gao2003} \\
  T-Cell Expected Kill Time & $10$ \textit{min} & Selected$^3$ \\
  Epithelial Cell Radius & $25 \mu m$ & Selected$^4$ \\
  T-Cell Age (at FOI) & $2$ \textit{hours} & Selected$^5$ \\
  T-Cell Age (in Blood) & $3$ \textit{days} & Selected$^5$ \\
  Onset of ATC Lymph Node Exit & $5$ \textit{days} & \cite{Banerjee2011} \\
  IgM Strength & Viral decay of $3$\textit{/day} & Selected$^6$ \\
  \hline  
\end{tabular}
\caption{Static model parameters.  1) Chosen to correspond to a 30 minute half-life.  2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.  This scales linearly with the number of virions in the cell's vicinity.  3) T cells induce apoptosis in nearby virus-secreting epithelial cells at a probabilistic rate such that the expected time to induce apoptosis is 10 minutes.  This rate does not scale with T cell numbers.  4)  The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns. 5) Parameters chosen after discussions with David Woodland, Trudeau Institute.  6) IgM presence is abstracted by increasing viral decay by a factor of three. }
\label{table:parameters}
\end{center}
\end{table}

\begin{table}
\centering
\begin{tabular}{ | r | c | c | c | }
  \hline                        
  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \footnotesize{(PFU/s$\cdot$cell)} \\
  \hline
  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & 5.4e-5 \\
   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \footnotesize{4.4e-5 --- 3.7e-4}\\ 
   \hline
  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & 3.8e-4 \\
   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} & \footnotesize{2.8e-4 --- 1.5e-3}\\
   \hline
  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & 5.1e-3 \\
   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} & \footnotesize{2.8e-3 --- 5.3e-3} \\
  \hline
\end{tabular}
\caption{Strain-specific model parameters.  Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.  Viral production values are taken from \cite{Mitchell2011}.}
\label{table:strains}
\end{table}

@@

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.

@@

Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells (Supplemental Methods).  Model fits, (Table \ref{table:strains}), were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There is no positive correlation between viral production and induced chemokine production across the three strains.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{data}
 \end{center}
\caption{Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.  sH1N1 IP-10 registered three values, not included in model fitting, above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).  Dashed lines show model fits to IP-10 and RANTES data.} 
 \label{fig:data}
\end{figure}

\subsection*{T-cell sensitivity to chemokine}

The model uses chemokine gradient as a surrogate for the inflammatory cytokines required for extravasation.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  However, T-cell sensitivity determines how strong the aggregate chemokine signal must be.

We simulated T cell sensitivity levels ranging over seven orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) between concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on T cell movement (model variance discussed in \textbf{S}2.1.).  Seasonal and pandemic H1N1 models simulated both IP-10 and RANTES; avian H5N1 included only RANTES (\textbf{S2.2}).  Because multiple T-cells in the same area do not enhance control of infection, we hypothesize more T cells are not better above a critical threshold.  We set the sensitivity to 10 $ng/mL$ (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.  Single runs were used due to computational limitations.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{sensitivity}
 \end{center}
\caption{Varying T-cell sensitivity to chemokine: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES. Total number of infected, expressing and apoptotic cells are plotted for each infection.  The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it. } 
 \label{fig:sensitivity}
\end{figure}

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T-cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T-cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  In aH5N1, the plaque is dense, allowing T-cells to find secreting cells easily, and infection is eliminated.  However, in both H1N1 simulations secreting cells were not eliminated.  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T-cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T-cells from completely clearing the infection.

T-cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T-cells can control (but not eliminate) the sH1N1 infection.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  That is, in each secreting cell a new viral particle is produced approximately every 200 seconds.  Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell could do is limit production to 18 new viral particles.  Thus, T-Cells alone cannot contain the pH1N1 infection.  In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{cycells}
 \end{center}
\caption{Simulated sH1N1 infection. Screenshots from day 4, day 5.5, and day 7.  The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells arriving at day 5 (green).  Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).} 
 \label{fig:cycells}
\end{figure}

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{plaquesize}
 \end{center}
\caption{Comparison of total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), and T cells (red) for the simulated infections of aH5N1, sH1N1, and pH1N1.  T cells are able to clear both secreting and incubating cells in aH5N1, are unable to contain incubating cells in sH1N1, and are unable to contain either type of infected cell in pH1N1.} 
 \label{fig:plaquesize}
\end{figure}

@@

%\subsection*{Caveats}

%Parameters are estimated to the best of our knowledge when values could not be found in literature.  We choose to ignore most aspects of the innate immune response to avoid over-parameterizing the model.  The unbounded growth of pH1N1 can partially be attributed to the lack of many innate mechanisms.  T cells are assumed to have the same level of sensitivity to both IP-10 and RANTES.  Furthermore, T cells climb chemokine gradient in straight lines as our experiments with biased random walks showed no effect on model behavior.

@@

Limitations in the model are primarily due to a number of simplifications to and deletions of elements in the innate immune response, allowing us to build a tractable model where data was available.  Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.  Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.  The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.  All of these rates may in fact be time-variable as indicated by superior data-fitting models \cite{Wu2011}.  Immigration and contributions of virus-nonspecific immune cells such as macrophages are not represented in our model.  Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection \cite{Miao2010}.  Thus, our model is not intended to predict clearance of virus from the lung.  Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.

The use of an ABM has certain advantages over a spatially homogeneous ODE model.  An ODE model assumes that any virus particle is capable of infecting any healthy cell.  Figure \ref{fig:cycells} shows that this is clearly not the case.  In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.  ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.  This is limiting as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.

>>>>

During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.  T cell navigation to the infected region is undirected but focused migration is made possible by local cytokine and chemokine production from infected epithelial cells.  We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using simulations in a purpose-built agent-based mathematical model (CyCells).

Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.  Of the chemokines tested, only secretion of CXCL10 (IP-10) and CCL5 (RANTES) was stimulated by infection with these three influenza strains, with IP-10 dominating the directed migration due to higher concentrations.  A delayed differential equation model was fit to the empirical per cell chemokine production rates, coupled with published T cell migration parameters, to calibrate the spatially explicit model in order to test inter-strain variation on T cell recruitment in the lungs.  We observe that T cell sensitivity to chemokine is just high enough to maximize their efficiency.  The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production and large plaque size.

The spatial nature of the model reveals unique challenges to T cell recruitment and motility not visible in spatially homogeneous ODE models.  Expanding plaque sizes isolate infected cells, impeding T cell discovery and infection control. In the model, T cells executed an efficient search with high sensitivity to either chemokine detected. A key limitation imposed on T cells was illustrated in their failure to clear the most rapidly replicating pandemic H1N1 virus after day 6, when T cells became ‘lost’ and inefficient in large infected foci as revealed in cinematic displays. This spatially explicit model is consistent with an efficient T cell recruitment to infected lung, without apparent means to induce increased efficiency, leaving vaccination as the key to shortening the interval between initial infection and T cell activation.

@@

Seasonal influenza infects up to 10,000,000 people and kills between 250,000 and 500,00 people every year.  Recent mutant strains of flu with high morbidity such as Avian H5N1 and 2009 Pandemic H1N1 raise the possibility of a global pandemic in the near future.  In order to clear an infection in the lung, T cells produced in the lymph nodes must navigate the branching bronchial vascular system in order to reach the site of infection.  Using an agent-based model, we look at how molecular signaling by cells infected with three differing strains of influenza aid this search process.  Our experimental data shows that two chemokines, IP-10 and RANTES, are help improve T cell search.  The spatial nature of our model reveals challenges to this search process not seen in conventional ordinary differential equation (ODE) models.  The simulated immune response is not able to contain the highly virulent 2009 pandemic H1N1 strain, suggesting that early intervention, i.e. vaccination, is key to containing the infection.

@@

Mathematical models have used detailed mouse data to characterize the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.  Such whole-response modeling has considerable potential to inform strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.  Models using ordinary differential equations to predict temporal events, however, do not account for spatial constraints, and must make assumptions about the efficiency of the viral spread and interaction of different populations.

Here we study the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).  To develop a model with tractable simulations, some components of host defense are abstracted and the model therefore does not aim to predict the true outcome for each strain examined.   We ask how very small foci of infected tissue can attract limited numbers of activated CD8 T cells after release into the systemic vascular system.  The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.   A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the local advantage brought by chemokine secretion in an inflammatory context have not been examined in detail in a virus-infected whole animal model.

@@

The standard hypothesis posits that the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.  Our model confirms this hypothesis, but we also observed three interesting characteristics of chemokine-directed migration not revealed by earlier data.  First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.  Second, each chemokine exhibited a clear threshold to direct migration, with diminishing improvements at increasing concentrations above the threshold.  Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.

@@

Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response. A simplified model of T cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions. The lung was represented as a two-dimensional sheet of healthy epithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T cells descend through the vascular tree until cytokine signal is detected on the local endothelium, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection. T cells that do not encounter cytokine recirculate to the lymph node. At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.

@@

In the model, epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}. \emph{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from {incubating} to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for approximately 17 hours and then die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T cells. Apoptoic cells continue to secrete virus until they die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T cells have three states. \emph{Circulating} T cells begin to emerge from the lymph node at five days post infection. \textit{Emigrating} T cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine. \emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{emigrating} and are reintroduced to a new random location in the lung. When a \emph{emigrating} T cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection. \emph{Chemotaxing} T cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells. \emph{Chemotaxing} cells decay exponentially with an average lifespan of two hours.

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \emph{expressing} epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration. Chemokine diffuses across the tissue but has no
direct effect beyond activating T cells. Both particle types decay exponentially.

@@

@@

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.  TNF was ignored as its effects are outside the scope of this paper.

@@

Chemokine data were collected from epithelial monolayers infected with three different influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..  To estimate per-cell production rates, we extended the ODE model of Ref. \cite{Mitchell2011} to represent chemokine production from infected cells (Eq. \textbf{S}1).  Model fits, (Table \ref{table:strains}), were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  The resulting chemokine production values were used in the CyCells ABM.  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There is no positive correlation between viral production and induced chemokine production across the three strains.

\subsection*{T cell sensitivity to chemokine}

The model uses chemokine gradient as a surrogate for the inflammatory cytokines required for extravasation.  Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.  However, T cell sensitivity determines how strong the aggregate chemokine signal must be.

We simulated T cell sensitivity levels ranging over seven orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) between concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on T cell movement (model variance discussed in \textbf{S}2.1.).  Seasonal and pandemic H1N1 models simulated both IP-10 and RANTES; avian H5N1 included only RANTES (\textbf{S2.2}).  Because multiple T cells in the same area do not enhance control of infection, we hypothesize more T cells are not more efficient above a critical threshold.  We set the sensitivity to 10 $ng/mL$ (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.  Single runs were used due to computational limitations.

@@

Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.  Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.  These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}). Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare. Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells. By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.  In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.  However, in both H1N1 simulations secreting cells were not eliminated.  Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..  T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.  Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.  Thus, T cells can fail to detect cellular changes in the plaque.  Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.

T cells are unable to kill cells that have not yet presented antigen.  At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}). By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.  The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.

A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.  That is, in each secreting cell a new viral particle is produced approximately every 200 seconds.  Assuming that secretion continues for one hour after apoptosis is initiated, the best a T cell could do is limit production to 18 new viral particles.  Thus, T cells alone cannot contain the pH1N1 infection.  In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T cells to limit a single infected cell to 1.3 viral particles in the one-hour window.  Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.

@@

@@

Limitations in the model are primarily due to a number of simplifications to and deletions of elements in the innate immune response, allowing us to build a tractable model where data was available.  Antigen presentation T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.  Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.  The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.  All of these rates may in fact be time-variable as indicated by superior data-fitting models \cite{Wu2011}.  Immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.  Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection \cite{Miao2010}.  Thus, our model is not intended to predict clearance of virus from the lung.  Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.

The use of an ABM has certain advantages over a spatially homogeneous ODE model.  An ODE model assumes that any virus particle is capable of infecting any healthy cell.  Figure \ref{fig:cycells} shows that this is clearly not the case.  In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.  ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.  This is limiting as can be seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.

Our ABM renders the model in OpenGL (Fig.~\ref{fig:cycells}).  Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems, for example, the spatial dynamics discussed earlier.  First, because T cells find infected cells by climbing a gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.   Second, T cell clustering persists after all virus-secreting cells have been eliminated.  The local chemokine maxima takes time to diffuse and decay so that T cells can climb the gradient to a new maximum.  Finally, we can see that infected cells are more disperse as infection size grows.  Because T cells are clustered they cannot cover the increasing plaque effectively.  These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.   The behavior of searching T cells described in this paper 
can enhance future global immune response models.

\subsection*{Chemokine Directed T Cell Search}
